Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model

### 2023-04-25 src/docs_21-06-23-mlgh/21-06-23-mlgh.Rmd
Update placeholder for MLGH presentation

<!-- # Bayesian modelling and inference -->

<!-- * As a statistical modeller, the bulk of our job is in constructing a generative model for data $y$ using parameters $\vartheta$ -->
<!-- * This is the joint distribution $p(y, \vartheta) = p(y \, | \, \vartheta) p(\vartheta)$ -->
<!-- * Then, we want to compute (ideally without having to think much about it) the posterior $p(\vartheta \, | \, y)$ which is \textcolor{hilit}{just}^[I've highlighted this with sarcasm in mind: it's a difficult problem] -->
<!-- $$ -->
<!-- p(\vartheta \, | \, y) = \frac{p(y, \vartheta)}{p(y)} = \frac{p(y \, | \, \vartheta) p(\vartheta)}{p(y)} -->
<!-- $$ -->
<!-- * The central problem of Bayesian inference is doing the following integral -->
<!-- $$ -->
<!-- p(y) = \int p(y, \vartheta) \text{d} \vartheta -->
<!-- $$ -->

<!-- # Numerical integration -->

<!-- * If you want to integrate something deterministically, you could use numerical integration, otherwise called \textcolor{hilit}{quadrature} -->
<!-- * Choose nodes $\vartheta \in \mathcal{Q} \subset \varTheta$ and weights $\omega: \varTheta \to \mathbb{R}$ and compute the weighted sum -->
<!-- $$ -->
<!-- \tilde p(y) = \sum_{\vartheta \in \mathcal{Q}} p(y, \vartheta) \omega(\vartheta) -->
<!-- $$ -->
<!-- * By "deterministic" I mean: if you follow the same procedure you will get the same answer -->

<!-- # Naive quadrature example -->

<!-- * Remember how integration is taught? (Rienmann sums) -->
<!-- * Try computing $\int_0^\pi \sin(x) \text{d}x = 2$ using trapezoid rule -->
<!-- * Nodes evenly spaced through $[0, \pi]$ -->

<!-- ```{r} -->
<!-- trapezoid_rule <- function(x, spacing) { -->
<!--   w <- rep(spacing, length(x)) # Weights given by space between nodes -->
<!--   w[1] <- w[1] / 2 # Apart from the first which is halved -->
<!--   w[length(x)] <- w[length(x)] / 2 # And the last, also halved -->
<!--   sum(w * x) # Compute the weighted sum -->
<!-- } -->
<!-- ``` -->

<!-- # -->

<!-- ```{r message=FALSE, echo=FALSE,  fig.cap="With 10 nodes it's 0.02 off."} -->
<!-- library(tidyverse) -->
<!-- cbpalette <- c("#56B4E9","#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999") -->

<!-- finegrid <- seq(0, pi, length.out = 1000) -->

<!-- plot <- data.frame(x = finegrid, y = sin(finegrid)) %>% -->
<!--   ggplot(aes(x = x, y = y)) + -->
<!--    geom_line(col = cbpalette[1]) + -->
<!--    theme_minimal() + -->
<!--    labs(x = "x", y = "sin(x)") -->

<!-- trapezoid_plot <- function(N) { -->
<!--   grid <- seq(0, pi, length.out = N) -->
<!--   int <- trapezoid_rule(x = sin(grid), spacing = grid[2] - grid[1]) -->

<!--   plot +  -->
<!--     geom_bar( -->
<!--       data = data.frame(x = grid, y = sin(grid)), -->
<!--       aes(x = x, y = y), alpha = 0.7, stat = "identity", -->
<!--       inherit.aes = FALSE, fill = cbpalette[2]) + -->
<!--     theme_minimal() + -->
<!--     labs( -->
<!--       subtitle = paste0("Number of nodes: ", N, "\nTrapezoid rule estimate: ", round(int, 3), "\nTruth: 2"), -->
<!--       x = "x", y = "sin(x)" -->
<!--     ) -->
<!-- } -->

<!-- trapezoid_plot(N = 10) -->
<!-- ``` -->

<!-- # -->

<!-- ```{r message=FALSE, echo=FALSE, fig.cap="With 30 nodes it's 0.002 off."} -->
<!-- trapezoid_plot(N = 30) -->
<!-- ``` -->

<!-- # -->

<!-- ```{r message=FALSE, echo=FALSE, fig.cap="With 100 nodes it's pretty much correct."} -->
<!-- trapezoid_plot(N = 100) -->
<!-- ``` -->

<!-- <!-- # Monte Carlo as it relates to numerical integration --> -->

<!-- <!-- * Suppose we can sample $\vartheta_i \sim p(y, \vartheta)$ for $i = 1, \ldots, N$ --> -->
<!-- <!-- * The Monte Carlo (MC) estimate is --> -->
<!-- <!-- $$ --> -->
<!-- <!-- \tilde p(y) = \frac{1}{N} \sum_i p(y, \vartheta_i) --> -->
<!-- <!-- $$ --> -->
<!-- <!-- which is quadrature with the samples as nodes and $\omega(\vartheta_i) = 1/N$ for all $i$ --> -->
<!-- <!-- * For complicated models^[Or not even that complicated] it's not possible to sample directly from $p(y, \vartheta)$, but we can usually sample from a Markov chain (MCMC) --> -->

<!-- <!-- # Monte Carlo is fundamentally unsound --> -->

<!-- <!-- * "Monte Carlo ignores information" according to @o1987monte --> -->
<!-- <!-- * Suppose $N = 3$ and we sample $\vartheta_1, \vartheta_2, \vartheta_3$ with $\vartheta_2 = \vartheta_3$ then our MC estimate is --> -->
<!-- <!-- $$ --> -->
<!-- <!-- \tilde p(y) = \frac{1}{3} \left( p(y, \vartheta_1) + p(y, \vartheta_2) + p(y, \vartheta_3) \right) --> -->
<!-- <!-- $$ --> -->
<!-- <!-- * This is despite the fact that nothing new about the function has been learned by adding $\{\vartheta_3, p(y, \vartheta_3)\}$ --> -->

<!-- <!-- # Application to HIV survey sampling --> -->

<!-- <!-- * This is a digression but... --> -->
<!-- <!-- * Say we're running a household survey, and sample the same individual twice --> -->
<!-- <!-- * We didn't learn anything new about HIV by surveying them again! --> -->
<!-- <!-- * This doesn't just bite for nodes or individuals which are exactly the same: an analogous argument can be made if they are close together and we expect their function evaluations to be similar --> -->

<!-- <!-- \begin{center} --> -->
<!-- <!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] --> -->
<!-- <!-- $\implies$ Bayesian quadrature, Bayesian survey design \\ --> -->
<!-- <!-- For some half-baked thoughts, see \texttt{athowes.github.io/fourth-gen/paper.pdf} --> -->
<!-- <!-- \end{tcolorbox} --> -->
<!-- <!-- \end{center} --> -->

<!-- # Curse of dimensionality -->

<!-- * Quadrature doesn't work very well when $\text{dim}(\vartheta)$ gets even moderately sized, because there is an exponential increase in the volume you need to cover -->
<!-- * If you're using $k$ points per dimension it's $k^{\text{dim}(\vartheta)}$ e.g. $k = 5$ then -->

<!-- ```{r} -->
<!-- 5^{1:8} -->
<!-- ``` -->

<!-- # Latent variables and hyperparameters -->

<!-- * Previously all of the parameters were under the symbol $\vartheta$ -- what if we split them up as being $\vartheta = (x, \theta)$ -->
<!-- * The key part about this is that $\dim(x) = N$ is big and $\dim(\theta) = m$ is small -->

<!-- | Names for $x$ | Names for $\theta$ | -->
<!-- |---------------|--------------------| -->
<!-- | Latent variables, random effects, latent field | Hyperparameters, fixed effects  -->

<!-- # Spatio-temporal statistics -->

<!-- * There is nothing inherently special about spatio-temporal statistics, but we do often end up tackling problems with similar structures -->
<!-- * We have observations indexed by space $s \in \mathcal{S}$ and time $t \in \mathcal{T}$ -->
<!-- * Usually we associate parameters to spatio-temporal locations, giving us something like $\{x_{s, t}\}_{s \in \mathcal{S}, t \in \mathcal{T}}$ -->

<!-- # What's important about this? -->

<!-- 1. There might be \textcolor{hilit}{a lot} of spatio-temporal locations, so $N$ might be pretty big! If you have 100 districts and 10 years, that's already $100 \times 10 = 1000$ parameters -->
<!-- 2. Perhaps we're willing to make quite strong assumptions about how things vary over space-time^[Are there any slides about spatial statistics that don't describe Tobler's first law of geography?]: not IID anymore! -->

<!-- <!-- As you get more data, the posterior approaches a Gaussian --> -->

<!-- # Latent Gaussian models -->

<!-- * A \textcolor{hilit}{latent Gaussian model} (LGM) [@rue2009approximate] looks along these lines: -->
<!-- \begin{alignat*}{2} -->
<!--   &\text{(Observations)}     &   y &\sim p(y \, | \, x, \theta), \\ -->
<!--   &\text{(Latent field)}     &   x &\sim \mathcal{N}(x  \, | \, \mu(\theta), Q(\theta)^{-1}), \\ -->
<!--   &\text{(Hyperparameters)}  &   \qquad \theta &\sim p(\theta). -->
<!-- \end{alignat*} -->
<!-- * Many models are LGMs, especially in spatio-temporal statistics -->
<!-- * $\dim(x) = n$, $\dim(x) = N$, $\dim(\theta) = m$ -->

<!-- # Laplace approximation -->

<!-- * Remember that we wanted to compute -->
<!-- $$ -->
<!-- p(y) = \int p(y, \vartheta) \text{d} \vartheta -->
<!-- $$ -->

<!-- * One trick for doing this is to pretend $p(\vartheta \, | \, y)$ is Gaussian -->

<!-- \begin{center} -->
<!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] -->
<!-- \begin{itemize} -->
<!-- \item Mode $\hat \vartheta = \argmax_\vartheta \log p(y, \vartheta)$ -->
<!-- \item Hessian $H(\hat \vartheta) = - \partial_\vartheta^2 \log p(y, \vartheta) \rvert_{\vartheta = \hat \vartheta}$ -->
<!-- \item Gaussian approximation $\implies \tilde p_{\texttt{G}}(\vartheta \, | \, y) = \mathcal{N}(\vartheta \, | \, \hat \vartheta, H(\hat \vartheta)^{-1})$ -->
<!-- \end{itemize} -->
<!-- \end{tcolorbox} -->
<!-- \end{center} -->

<!-- # Example of computing the Laplace approximation -->

<!-- * Consider the following model for $i = 1, \ldots, n$ with fixed $a$ and $b$ -->
<!-- $$ -->
<!-- y_i \sim \text{Poisson}(\lambda), \quad -->
<!-- \lambda \sim \text{Gamma}(a, b). -->
<!-- $$ -->
<!-- * It's conjugate so we directly know that -->
<!-- $$ -->
<!-- \lambda \, | \, y \sim \text{Gamma}(a + \sum_i y_i, b + n) -->
<!-- $$ -->

<!-- # -->

<!-- ```{r message=FALSE, echo=FALSE,  fig.cap="A Gamma prior with $a = 3$ and $b = 1$."} -->
<!-- a <- 3 -->
<!-- b <- 1 -->

<!-- prior <- ggplot(data = data.frame(x = c(0, 10)), aes(x)) + -->
<!--   stat_function(fun = dgamma, n = 500, args = list(shape = a, rate = b), col = cbpalette[1]) + -->
<!--   annotate("text", x = 6, y = 0.15, label = "Gamma(3, 1)", col = cbpalette[1], size = 5) + -->
<!--   labs(x = "", y = "") + -->
<!--   theme_minimal() -->

<!-- prior -->
<!-- ``` -->

<!-- # -->

<!-- ```{r message=FALSE, echo=FALSE,  fig.cap="Draw 3 points from $\\text{Poisson}(3)$, then compute the posterior."} -->
<!-- set.seed(2) -->
<!-- y <- rpois(3, lambda = 2) -->

<!-- posterior <- prior + -->
<!--   geom_point(data = data.frame(x = y, y = 0), aes(x = x, y = y), inherit.aes = FALSE, alpha = 0.7, size = 2) + -->
<!--   stat_function(data = data.frame(x = c(0, 10)), aes(x), fun = dgamma, n = 500, args = list(shape = a + sum(y), rate = b + length(y)), col = cbpalette[2]) + -->
<!--   annotate("text", x = 5, y = 0.25, label = "Gamma(9, 4)", col = cbpalette[2], size = 5) -->

<!-- posterior -->
<!-- ``` -->

<!-- # -->

<!-- ```{r message=FALSE} -->
<!-- fn <- function(x) dgamma(x, a + sum(y), b + length(y), log = TRUE) -->

<!-- # Here we are using numerical derivatives rather than automatic -->
<!-- ff <- list( -->
<!--   fn = fn, -->
<!--   gr = function(x) numDeriv::grad(fn, x), -->
<!--   he = function(x) numDeriv::hessian(fn, x) -->
<!-- ) -->

<!-- opt_bfgs <- aghq::optimize_theta( -->
<!--   ff, 1, control = aghq::default_control(method = "BFGS") -->
<!-- ) -->
<!-- ``` -->

<!-- # Laplace approximation -->

<!-- ```{r} -->
<!-- laplace <- posterior + -->
<!--   stat_function( -->
<!--     data = data.frame(x = c(0, 10)), -->
<!--     aes(x), -->
<!--     fun = dnorm, -->
<!--     n = 500, -->
<!--     args = list(mean = opt_bfgs$mode, sd = sqrt(1 / opt_bfgs$hessian)), -->
<!--     col = cbpalette[3] -->
<!--   ) -->
<!-- ``` -->

<!-- # Laplace approximation -->

<!-- ```{r message=FALSE, echo=FALSE,  fig.cap="The Laplace approximation is good near the mode but not in the tails."} -->
<!-- laplace + -->
<!--   annotate("text", x = 5.5, y = 0.35, label = "Laplace approximation", col = cbpalette[3], size = 5) -->
<!-- ``` -->

<!-- # Laplace approximation -->

<!-- * Now -->
<!-- $$ -->
<!-- p(y) = \frac{p(\vartheta, y)}{p(\vartheta \, | \, y)} \approx \frac{p(\vartheta, y)}{p_{\texttt{G}}(\vartheta \, | \, y)} -->
<!-- $$ -->
<!-- and we can evaluate RHS where we would like, so let's pick the point at which the Gaussian is most accurate, which is $\hat \vartheta$ -->
<!-- $$ -->
<!-- p_{\texttt{LA}}(y) = \frac{p(\vartheta, y)}{p_{\texttt{G}}(\vartheta \, | \, y)} \rvert_{\vartheta = \hat \vartheta} = (2 \pi)^{\dim(\vartheta) / 2} \det(H(\hat \vartheta))^{- 1 / 2} p(\hat \vartheta, y) -->
<!-- $$ -->

<!-- # Marginal Laplace approximation -->

<!-- * Is it reasonable to just assume $p(\vartheta \, | \, y)$ is Gaussian? -->
<!-- * There is the Bernstein–von Mises theorem, but we can't rely on that in general. However... -->

<!-- 1. We just described a class of models (LGMs) where some subset of the parameters ($x$) have a Gaussian prior $\implies$ it's a lot more reasonable to think that they would have a marginal posterior which is close to Gaussian -->
<!-- 2. We just talked about how big $x$ is in comparison to $\theta$! $\implies$ most of the work in our integral can be done using a \textcolor{hilit}{marginal Laplace} approximation to get rid of $x$ -->

<!-- # Generalist versus specialist methods -->

<!-- Dichotomy in statistical inference methods: -->

<!-- \begin{center} -->
<!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] -->

<!-- \begin{enumerate} -->
<!--   \item Generalist: works in all situations -->
<!--   \item Specialist: "exploits" properties of the problem at hand -->
<!-- \end{enumerate} -->

<!-- \end{tcolorbox} -->
<!-- \end{center} -->

<!-- We are taking approach 2! -->

<!-- # Marginal Laplace approximation -->

<!-- * What does this look like? Instead of assuming $p(\vartheta \, | \, y) = p(x, \theta \, | \, y)$ is Gaussian we assume $p(x \, | \, \theta, y)$ is -->
<!-- $$ -->
<!-- \tilde p_{\texttt{G}}(x \, | \, \theta, y) = \mathcal{N}(x \, | \, \hat x, H(\hat x))^{-1} -->
<!-- $$ -->
<!-- where $\hat x = \hat x(\theta)$ -->
<!-- * Now the marginal Laplace approximation is -->
<!-- $$ -->
<!-- p_{\texttt{LA}}(\theta, y) = \frac{p(x, \theta, y)}{\tilde p_{\texttt{G}}(x \, | \, \theta, y)} \rvert_{x = \hat x} = (2 \pi)^{N / 2} \det(H(\hat x))^{- 1 / 2} p(\hat x, \theta, y) -->
<!-- $$ -->

<!-- # Integrated nested Laplace approximation -->

<!-- * Now we can compute $p_{\texttt{LA}}(\theta, y)$ -- but what we really want is still $p(y)$ for which we need to integrate out $\theta$ -->
<!-- * Hopefully^[Really: hopefully] $m$ is small enough that we can now tackle this with quadrature -->
<!-- * Given nodes $\mathcal{Q}$ and a weighting function $\omega$ we can compute -->
<!-- $$ -->
<!-- p(y) \approx \sum_{\theta \in \mathcal{Q}} p_{\texttt{LA}}(\theta, y) \omega(\theta) -->
<!-- $$ -->
<!-- * This is the famous integrated nested Laplace approximation (INLA) -->

<!-- # Taking stock -->

<!-- 1. Bayesian inference is integration -->
<!-- 2. Spatial statistics has parameters $(x, \theta)$ -->
<!-- 3. Integrate $x$ cheaply using a Gaussian assumption -->
<!-- 4. Try a bit harder with $\theta$ performing quadrature -->

<!-- \begin{center} -->
<!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] -->

<!-- Now let's apply this to a difficult problem in HIV inference! -->

<!-- \end{tcolorbox} -->
<!-- \end{center} -->

<!-- # The Naomi model -->

<!-- :::::::::::::: {.columns} -->

<!-- ::: {.column width=.65} -->

<!-- * Naomi is a spatial evidence synthesis model -->
<!-- * Used by countries to produce HIV estimates in a yearly process supported by UNAIDS -->
<!-- * Inference for Naomi is currently conducted using `TMB` by optimising $p_{\texttt{LA}}(\theta, y)$, and has to be pretty quick to allow for interactive review and development of estimates -->

<!-- ::: -->

<!-- ::: {.column width=.35} -->

<!-- ```{r, echo=FALSE, fig.cap="A supermodel", out.width = '65%'} -->
<!-- knitr::include_graphics("naomi_hex.png") -->
<!-- ``` -->

<!-- ::: -->

<!-- :::::::::::::: -->

<!-- #  -->

<!-- ```{r, echo=FALSE, fig.cap="Example of the user interface from \\texttt{https://naomi.unaids.org/}", out.width = '70%'} -->
<!-- knitr::include_graphics("naomi_user.png") -->
<!-- ``` -->

<!-- # Template Model Builder refresher -->

<!-- * To use `TMB` [@kristensen2015tmb], you write your objective function $- \log p(y, x, \theta)$ in the C++ syntax -->
<!-- * For example, for the model $\y \sim \mathcal{N}(\mu, 1)$ with $p(\mu) \propto 1$ then the `TMB` user template looks like (next slide) -->
<!-- * `TMB` implements the Laplace approximation! Set `random = "x"` -->

<!-- #  -->

<!-- ```{cpp eval=FALSE} -->
<!-- #include <TMB.hpp> -->

<!-- template <class Type> -->
<!-- Type objective_function<Type>::operator()() { -->
<!--   // Define data e.g. -->
<!--   DATA_VECTOR(y); -->
<!--   // Define parameters e.g. -->
<!--   PARAMETER(mu); -->
<!--   // Calculate negative log-likelihood e.g. -->
<!--   nll = Type(0.0); -->
<!--   nll -= dnorm(y, mu, 1, true).sum() -->
<!--   return(nll); -->
<!-- } -->
<!-- ``` -->

<!-- # Why do we use `TMB` -->

<!-- * It runs quickly, and is flexible enough to write the model -->
<!-- * Another answer is that we don't have any better options -->

<!-- | Option | Would it work? | -->
<!-- |--------|--------------------| -->
<!-- | MCMC | No! It is accurate (eventually) but takes too long | -->
<!-- | INLA via `R-INLA`        | No! Though Naomi is a spatial model with a large Gaussian latent field, it isn't technically a LGM, and isn't compatible with `R-INLA` | -->

<!-- # Idea -->

<!-- * Implement an algorithm inspired by INLA which -->
<!-- 1. is compatible with Naomi and its intricacies -->
<!-- 2. uses `TMB` -- and thereby automatic differentiation -- to perform the Laplace approximation -->

<!-- # Not a new idea! -->

<!-- > My main comment is that several aspects of the computational machineery that is presented by Rue and his colleagues **could benefit from the use of a numerical technique known as automatic differentiation (AD)** ... By the use of AD one could obtain a system that is automatic from a user's perspective... the benefit would be a fast, flexible and easy-to-use system for doing Bayesian analysis in models with Gaussian latent variables  -->

<!-- -- Hans J. Skaug (coauthor of `TMB`), RSS discussion of @rue2009approximate -->

<!-- # Meanwhile in Canada... -->

<!-- * Alex Stringer in Toronto $\to$ Waterloo had independently been thinking along similar lines, and made a lot of progress -->

<!-- 1. Implementing an algorithm similar to INLA, using a specific quadrature rule $\mathcal{Q}$ called adaptive Gauss-Hermite quadrature (AGHQ) which arguably should be the default for this problem -->
<!-- 2. Defining a class of models called extended latent Gaussian models (ELGMs) which Naomi fits into -->

<!-- * I will explain what both of these acronyms (AGHQ^[Amusingly similar to AGYW: adolescent girls and young women] and ELGM) mean! -->

<!-- # Gauss-Hermite quadrature -->

<!-- * Recall -->
<!-- $$ -->
<!-- \int f(z) \text{d} z \approx \sum_{z \in \mathcal{Q}} f(z) \omega(z) -->
<!-- $$ -->
<!-- * Replace $f(z)$ by $\phi(z) f(z)$ and say that $f$ is a polynomial and $\phi$ is unknown -->
<!-- * Suppose that $\phi(z) = \exp(-z^2)$ -->

<!-- # Adaptation -->

<!-- * The nodes and weights we use should probably depend on the integrand -->
<!-- * Especially when the integrand is also a function of $y$, which we don't know in advance, as in $p(y, \vartheta)$ -->
<!--   * i.e. how could any fixed quadrature rule be appropriate for all possible $y$? -->
<!-- * Let $z \in \mathcal{Q}(m, k)$ then -->
<!-- $$ -->
<!-- \theta(z) = \hat \theta + Lz -->
<!-- $$ -->
<!-- where $L$ is the lower Cholesky of $H = LL^\top$ -->
<!-- * Other matrix decompositions can also be used e.g. spectral $H = E \Lambda E^\top = (E\Lambda^{1/2})(E\Lambda^{1/2})$ -->
<!--   * Arguably this is preferable: symmetric with respect to the principle axis [@jackel2005note] -->

<!-- #  -->

<!-- ```{r, echo=FALSE, message=FALSE, fig.cap="Unadapted points in two dimensions with $k = 3$."} -->
<!-- mu <- c(1, 1.5) -->
<!-- cov <- matrix(c(2, 1, 1, 1), ncol = 2) -->

<!-- obj <- function(theta) { -->
<!--   mvtnorm::dmvnorm(theta, mean = mu, sigma = cov) -->
<!-- } -->

<!-- grid <- expand.grid( -->
<!--   theta1 = seq(-2, 5, length.out = 700), -->
<!--   theta2 = seq(-2, 5, length.out = 700) -->
<!-- ) -->

<!-- ground_truth <- cbind(grid, pdf = obj(grid)) -->

<!-- plot0 <- ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) + -->
<!--   geom_contour(col = cbpalette[1]) + -->
<!--   coord_fixed(xlim = c(-2, 4.5), ylim = c(-2, 4.5), ratio = 1) + -->
<!--   labs(x = "", y = "") + -->
<!--   theme_minimal() -->

<!-- gg <- mvQuad::createNIGrid(2, "GHe", 3) -->

<!-- add_points <- function(plot0, gg) { -->
<!--   plot0 + -->
<!--     geom_point( -->
<!--       data = mvQuad::getNodes(gg) %>% -->
<!--               as.data.frame() %>% -->
<!--               mutate(weights = mvQuad::getWeights(gg)),  -->
<!--       aes(x = V1, y = V2, size = weights), -->
<!--       alpha = 0.8, -->
<!--       col = cbpalette[2], -->
<!--       inherit.aes = FALSE -->
<!--     ) + -->
<!--     scale_size_continuous(range = c(1, 2)) -->
<!-- } -->

<!-- add_points(plot0, gg) + -->
<!--   labs(size = "Weight", caption = "") -->
<!-- ``` -->

<!-- # -->

<!-- ```{r, echo=FALSE, message=FALSE, fig.cap="Add the mean $z + \\hat \\theta$."} -->
<!-- gg2 <- gg -->
<!-- mvQuad::rescale(gg2, m = mu, C = diag(c(1, 1)), dec.type = 2) -->

<!-- add_points(plot0, gg2) + -->
<!--   labs(size = "Weight", caption = "") -->
<!-- ``` -->

<!-- # -->

<!-- ```{r, echo=FALSE, message=FALSE, fig.cap="First option: rotate by the lower Cholesky $Lz + \\hat \\theta$."} -->
<!-- gg3 <- gg -->
<!-- mvQuad::rescale(gg3, m = mu, C = cov, dec.type = 2) -->

<!-- add_points(plot0, gg3) + -->
<!--   labs(size = "Weight", caption = "") -->
<!-- ``` -->

<!-- # -->

<!-- ```{r, echo=FALSE, message=FALSE, fig.cap="Second option: rotate using the eigendecomposition $E \\Lambda^{1/2} z + \\hat \\theta$."} -->
<!-- gg3 <- gg -->
<!-- mvQuad::rescale(gg3, m = mu, C = cov, dec.type = 1) -->

<!-- add_points(plot0, gg3) + -->
<!--   labs(size = "Weight", caption = "") -->
<!-- ``` -->

<!-- # Extended latent Gaussian models -->

<!-- * In an LGM, the conditional mean depends on exactly one structured additive predictor -->
<!-- \begin{align*} -->
<!-- y_i &\sim p(y_i \, | \, \eta_i, \theta_1), \quad i \in [n]\\ -->
<!-- \mu_i &= \mathbb{E}(y_i \, | \, \eta_i) = g(\eta_i), \\ -->
<!-- \eta_i &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}), -->
<!-- \end{align*} -->
<!-- where $\beta_0$, $\{\beta_j\}$ and $\{f_k(\cdot)\}$ have Gaussian priors (and can be collected into the latent field $x$) -->

<!-- # Extended latent Gaussian models -->

<!-- * ELGM remove this requirement such that -->
<!-- $$ -->
<!-- \mu_i = g_i(\eta_{\mathcal{J}_i}) -->
<!-- $$ -->
<!-- where $g_i: \mathbb{R}^{|\mathcal{J}_i|} \to \mathbb{R}$ and $\mathcal{J}_i$ is some set of indices -->
<!-- * Let $\dim(\eta) = N_n$ with $\mathcal{J}_i \subseteq \{1, \ldots, N_n\}$ -->
<!--   * $N_n < n$: more data points than structured additive predictors -->
<!--   * $N_n = n$: as many data points as structured additive predictors (LGM case) -->
<!--   * $N_n > n$: fewer data points than structured additive predictors -->
<!-- * The $g_i$ allow for a higher degree of non-linearity in the model -->

<!-- # Why is Naomi an ELGM? -->

<!-- | | Things I thought made Naomi an ELGM | Does it? | -->
<!-- |-|---------------|--------------------| -->
<!-- |1| ANC offset from household survey | ? | -->
<!-- |2| Incidence depends on adult prevalence and coverage | ? | -->
<!-- |3| ART coverage and recent infection are products | ? | -->
<!-- |4| ART attendance uses a multinomial | ? | -->
<!-- |5| Aggregation of finer processes | ? | -->
<!-- |6| Multiple link functions | ? | -->

<!-- * I will explain each of these^[The notation may not have been introduced properly, but hopefully the gist will still make sense], and whether or not they make Naomi an ELGM, in more detail in slides to follow -->

<!-- # ANC offset from household survey -->

<!-- * Linear predictors for ANC indicators contain nested in them the linear predictors for household survey indicators -->
<!-- \begin{align*} -->
<!-- \text{logit}(\rho_{x, a}^{\text{ANC}}) &= \text{logit}(\rho_{x, F, a}) + \beta^{\rho^{\text{ANC}}} + u_x^{\rho^{\text{ANC}}} + \eta_{R_x, a}^{\rho^{\text{ANC}}}, \\ -->
<!-- \text{logit}(\alpha_{x, a}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a}) + \beta^{\alpha^{\text{ANC}}} + u_x^{\alpha^{\text{ANC}}} + \eta_{R_x, a}^{\alpha^{\text{ANC}}}.  -->
<!-- \end{align*} -->
<!-- * Here $\text{logit}(\rho_{x, F, a})$ and $\text{logit}(\alpha_{x, F, a})$ *are* Gaussian, but we have dependency of $\mu_i$ on two $\eta_i$ -->
<!-- * Conclusion: \textcolor{hilit}{does} make Naomi an ELGM -->

<!-- # ANC offset from household survey -->

<!-- * Note that `R-INLA` does have the `copy` feature $\eta^\star = A \eta$ where $A$ is $n \times n$^[I've also seen it claimed that it could be $m \times n$ where $m \neq n$, so I'm unsure which is right] -->
<!-- $$ -->
<!-- \begin{pmatrix} -->
<!-- \eta_1^\star \\ -->
<!-- \eta_2^\star -->
<!-- \end{pmatrix} -->
<!-- = -->
<!-- \begin{pmatrix} -->
<!-- 1 & 1 \\ -->
<!-- 0 & 1 -->
<!-- \end{pmatrix} -->
<!-- \begin{pmatrix} -->
<!-- \eta_1 \\ -->
<!-- \eta_2 -->
<!-- \end{pmatrix} -->
<!-- = -->
<!-- \begin{pmatrix} -->
<!-- \eta_1 + \eta_2 \\ -->
<!-- \eta_2 -->
<!-- \end{pmatrix} -->
<!-- $$ -->
<!-- * By having effects only apply to a subset of indices (`idx = c(NA, 1)` say) perhaps it can work -->

<!-- \begin{center} -->
<!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] -->

<!-- Warning! -->

<!-- \begin{enumerate} -->
<!--     \item Is an LGM $\nRightarrow$ can be fit with \texttt{R-INLA} -->
<!--     \item Can be fit with \texttt{R-INLA} $\nRightarrow$ is an LGM -->
<!-- \end{enumerate} -->

<!-- \end{tcolorbox} -->
<!-- \end{center} -->

<!-- # Incidence depends on adult prevalence and coverage -->

<!-- * Linear predictor for incidence contains aggregated prevalence and coverage -->
<!-- $$ -->
<!-- \log(\lambda_{x, s, a}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x}^{\text{15-49}}) + u_x^\lambda + \eta_{R_x, s, a}^\lambda. -->
<!-- $$ -->
<!-- * Here $\log(\rho_{x}^{\text{15-49}})$ and $\log(1 - \omega \cdot \alpha_{x}^{\text{15-49}})$ are not going to be Gaussian -->
<!-- * Conclusion: \textcolor{hilit}{does} make Naomi an ELGM -->

<!-- # ART coverage and recent infection are products -->

<!-- * In the household survey, say, individuals who are taking ART or have been recently infected must be HIV positive -->
<!-- \begin{align*} -->
<!-- y^{\hat \alpha}_{x, s, a} &\sim \text{xBin}(m_{x, s, a}, \rho_{x, s, a} \cdot \alpha_{x, s, a}), \\ -->
<!-- y^{\hat \kappa}_{x, s, a} &\sim \text{xBin}(m_{x, s, a}, \rho_{x, s, a} \cdot \kappa_{x, s, a}). -->
<!-- \end{align*} -->
<!-- * $\text{logit}(\rho_{x, s, a})$ and $\text{logit}(\alpha_{x, s, a})$ are Gaussian, but we're taking a product here -->
<!-- * $\kappa_{x, s, a}$ is more complicated: a function of incidence, prevalence, mean duration of recent infection and false recent ratio -->
<!-- * Conclusion: \textcolor{hilit}{does} make Naomi an ELGM -->

<!-- \begin{center} -->
<!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] -->
<!-- Warning! These equations as writen do not appear in Naomi: instead there are aggregated versions, more about this soon. -->
<!-- \end{tcolorbox} -->
<!-- \end{center} -->

<!-- # ART attendance uses the multinomial -->

<!-- # Aggregation of finer processes -->

<!-- * There are many instances of models being placed on aggregate quantities -->
<!-- \begin{align*} -->
<!-- y^{\hat \theta}_{\mathcal{I}} &\sim \text{xBin}(m^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}}), \\ -->
<!-- \rho_{\mathcal{I}} &= \frac{\sum_{i \in \mathcal{I}} N_i \rho_i}{\sum_{i \in \mathcal{I}} N_i}. -->
<!-- \end{align*} -->
<!-- * Here we have $|\mathcal{I}|$ linear predictors being informed by one observation -->
<!-- * Conclusion: \textcolor{hilit}{does} make Naomi an ELGM -->

<!-- \begin{center} -->
<!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] -->
<!-- I've previously worked on this "disaggregation regression" situation with respect to space, see \texttt{athowes/areal-comparison} -->
<!-- \end{tcolorbox} -->
<!-- \end{center} -->

<!-- # Multiple link functions -->

>>>>

# The Naomi model

:::::::::::::: {.columns}

::: {.column width=.65}

* Naomi is a complicated spatio-temporal evidence synthesis model
* Used by countries to produce HIV estimates in a yearly process supported by UNAIDS
* Fast inference is important to allow for interactive review and development of estimates
* Inference for Naomi is currently conducted using Template Model Builder (`TMB`) [@kristensen2016tmb]

:::

::: {.column width=.35}

```{r, echo=FALSE, fig.cap="A supermodel", out.width = '65%'}
knitr::include_graphics("naomi_hex.png")
```

:::

::::::::::::::

#

```{r, echo=FALSE, fig.cap="Example of the user interface from \\texttt{https://naomi.unaids.org/}", out.width = '70%'}
knitr::include_graphics("naomi_user.png")
```

# Thanks for listening!

* Working on a paper based on this work called "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model" joint with Alex Stringer (Waterloo), Seth Flaxman (Oxford), Jeff Eaton (Imperial)
* Code and notebooks for this project are available at `athowes.github.io/elgm-inf`

\begin{center}
\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]

Let me know if you'd be up for being an early reader!

\end{tcolorbox}
\end{center}
